---
title: "Causal Inference Methods"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    source: embed
runtime: shiny
---





```{r,warning=FALSE,message=FALSE}

library(tidyverse)
library(plm)
library(DT)
library(stargazer)
library(shiny)
library(RCurl)
library(plotly)
library(broom)
library(shiny)
library(flexdashboard)
library(MatchIt)
library(shinyWidgets)
library(halfmoon)
library(sandwich)
library(fixest)
library(tidysynth)



# FIXED-EFFECTS: TAB 1  
  
# DATA PROCESSING AND READING FOR CAUSAL INFERENCE METHODS

rm(list = ls())
## FIXED-EFFECTS

data(Crime)
df<- Crime 

dataCrime<-filter(df,county %in% c(1,3,7,23))



dataCrime_within<-dataCrime %>%
  group_by(county) %>%
  mutate(crime_within = crmrte - mean(crmrte),
         prbarr_within = prbarr - mean(prbarr))


### LM MODEL WITH POOLED VS FIXED-EFFECTS


pooled<-tidy(feols(crmrte ~ prbarr, dataCrime))
fe<-tidy(feols(crmrte ~ prbarr | county, data = dataCrime ))



# MATCHING: TAB 2


# DATA PROCESSING


url<-"https://github.com/rudymeza54/causal_inference/blob/main/mosquito_nets.RDS?raw=true"


malaria_df <- readRDS(gzcon(url(url)))


malaria_df$net<-as.factor(malaria_df$net)



# MATCHIT DATA PROCESSING



m1 = matchit(net ~ income + temperature + health,
             method = "nearest", distance = "mahalanobis",
             replace = FALSE,data = malaria_df)

m1data = match.data(m1)


# REGRESSION MODELS

naive<-tidy(lm(malaria_risk ~ net,malaria_df)) 
matchit<-tidy(lm(malaria_risk ~ net,m1data))

# IPW MODELING WITH ROBUST STANDARD ERRORS

model_logit <- glm(net ~ income + temperature + health,
                   data = malaria_df,
                   family = binomial(link = "logit"))

net_probabilities <- augment_columns(model_logit,
                                     malaria_df,
                                     type.predict = "response") |>
  # The predictions are in a column named ".fitted", so we rename it here
  rename(propensity = .fitted)


net_ipw <- net_probabilities |>
  mutate(ipw = (net_num / propensity) + ((1 - net_num) / (1 - propensity)))



model_ipw <- lm(malaria_risk ~ net,
                data = net_ipw,
                weights = ipw)

robust_var <- sandwich(model_ipw)[2, 2]
point_est <- coef(model_ipw)[2]
lb <- point_est - 1.96 * sqrt(robust_var)
ub <- point_est + 1.96 * sqrt(robust_var)

main_output <-tibble(
        net = point_est,
        lower = lb,
        upper = ub,
        significant = ifelse((lb >0 & ub> 0) | (lb < 0 & ub < 0),"Significant","Not Significant")
      )
# SYNTHETIC CONTROL: SMOKING DATASET AND IPW WITH DIFF-IN-DIFF: TAB 3

pre<-smoking %>%
  filter(
         year <1988)%>%
  mutate(group = ifelse(state =="California",1,0))

post<-smoking %>%
  filter(
         year >= 1988)%>%
  mutate(group = ifelse(state =="California",1,0),
         ipw = 1)



model_logit_smoking <- glm(group ~ retprice + beer + age15to24,
                   data = pre,
                   family = binomial(link = "logit"))

smoking_probabilities <- augment_columns(model_logit_smoking,
                                     pre,
                                     type.predict = "response") |>
  # The predictions are in a column named ".fitted", so we rename it here
  rename(propensity = .fitted)


smoking_ipw <- smoking_probabilities |>
  mutate(ipw = (group / propensity) + ((1 - group) / (1 - propensity)))%>%
  select(-c(".se.fit",".resid",".hat",".sigma",".cooksd",".std.resid"))




main<-smoking_ipw %>%
  bind_rows(post)%>%
  mutate(treated = state=="California",
           post = year > 1988)


 x_cig<- lm(cigsale ~ treated*post, main,weights = ipw)
      
      robust_var <- sandwich(x_cig)[4,4]
      point_est <- coef(x_cig)[4]
      lb <- point_est - 1.96 * sqrt(robust_var)
      ub <- point_est + 1.96 * sqrt(robust_var)
      

      
model_output_cig <-tibble(
        var = "California",
        ate = point_est,
        lower = lb,
        upper = ub,
        significant = ifelse((lb >0 & ub> 0) | (lb < 0 & ub < 0) 
                             ,"Significant","Not Significant"))


# PLOTS FOR DIFF-IN-DIFF



plot1<-smoking_ipw %>%
  group_by(year,group)%>%
  summarize(cigs = sum(cigsale*ipw)/sum(ipw))

plot2<-post %>%
  group_by(year,group)%>%
  summarize(cigs = sum(cigsale*ipw)/sum(ipw))


main_plot<-bind_rows(plot1,plot2)





```






<b>Fixed-Effects</b> {data-orientation=rows}
=====================================  


Sidebar {.sidebar}
-------------------------------------

```{r}

radioButtons("fe",label = "Select Pooled or Fixed-Effects:",
choices = c("Pooled Model","Fixed-Effects Model"),selected = "Pooled Model")





```



<h3 class="header">Pooled Model vs Fixed-Effects Model</h3>

<br>

<p1 class="p1">
Looking at Crime Rate ~ Probability of Arrest

Pooling Bias-The naive model groups together different counties

We want to compare the county to itself (Within)
</p1>






Column {data-width=650} 
-------------------------------------

### <b>Crime Rate and Probability of Arrest</b>

```{r}


renderPlotly({
  if(input$fe=="Pooled Model"){
    
  ggplot(dataCrime) +
  aes(x = prbarr, y = crmrte, colour = county) +
  geom_jitter(size = 2.4) +
  scale_color_viridis_c(option = "cividis", direction = 1) +
  theme_minimal()+
  geom_smooth(method = lm, formula = y ~ x)->p
    
  }
  else if(input$fe=="Fixed-Effects Model"){
    
  ggplot(dataCrime_within) +
  aes(x = prbarr_within, y = crime_within, colour = county) +
  geom_jitter(size = 2.4) +
  scale_color_viridis_c(option = "cividis", direction = 1) +
  theme_minimal()+
  geom_smooth(method = lm, formula = y ~ x)->p
    
  }
  ggplotly(p)
})



```







Row {data-width=650} 
-------------------------------------

### <b>Model Output</b>

```{r}

renderDT({
  if(input$fe=="Pooled Model"){
    
    pooled
  }
  else if(input$fe=="Fixed-Effects Model"){
    
    fe
  }
  
})



```




<b>IPW (Matching)</b> {data-orientation=rows}
===================================== 





Sidebar {.sidebar}
-------------------------------------

```{r}

radioButtons("ipw",label = "Select Model:",
            choices = c("Naive Model","MatchIT","IPW Model"),selected = "Naive Model")


materialSwitch(inputId = "id", label = "Table switch", status = "primary")




```


<h3 class="header">Naive, MatchIt and IPW Matching</h3>

<p1 class="p1">
This is a simulated dataset. 

The ATE is <b>-10</b>

Naive Model: Malaria Risk ~ netTRUE

MatchIT: Malaria Risk ~ netTRUE<br>

nearest neighbor matching based on variables temperature, health, and income

IPW Matching: Malaria Risk ~ netTRUE<br>

Inverse Probability Weight matching based off variables temperature, health, and income.

Assumptions: Conditional Independence Assumption ~ we've closed all backdoors with our covariates.
</p1>

Column
-------------------------------------

### <b>Model</b>

```{r}

renderValueBox({
  
  if(input$ipw=="Naive Model"){
    valueBox(naive$term[2])->t
  } else if(input$ipw=="MatchIT"){
    valueBox(matchit$term[2])->t
  
  } else if(input$ipw=="IPW Model"){
    valueBox("netTRUE")->t
  
}
  
  
  t
  
})




```



### <b>Average Treatment Effect (ATE)</b>

```{r}


renderValueBox({
  
  if(input$ipw=="Naive Model"){
    valueBox(round(naive$estimate[2],2))->x
  } else if(input$ipw=="MatchIT"){
    valueBox(round(matchit$estimate[2],2))->x
  
  }else if(input$ipw=="IPW Model"){
    valueBox(round(main_output$net,2))->x
  
}
  x
})

```




### <b>P-Value</b>

```{r}

renderValueBox({
  
  if(input$ipw=="Naive Model"){
  
  final<-naive %>%
    mutate(
      p.value = ifelse(p.value < .05,"significant",
                       "not significant")
    )
  
  valueBox(final$p.value[2])->y
  
  } else if(input$ipw=="MatchIT"){
    
    
    final<-matchit %>%
    mutate(
      p.value = ifelse(p.value < .05,"significant",
                       "not significant")
    )
    
    valueBox(final$p.value[2])->y
    
    
  } else if(input$ipw=="IPW Model"){
    
    valueBox(main_output$significant)->y
  }

  y
  
})


```


Row {data-width=650} 
-------------------------------------


### <b>Malaria Risk</b>

```{r}

uiOutput("dynamic")

output$dynamic <- renderUI({ 
  if(input$ipw=="Naive Model"){
    
    if (input$id == "FALSE") plotlyOutput("plot")
    
    else if (input$id == "TRUE") dataTableOutput("table")
  }
  else if(input$ipw=="MatchIT"){
    
    if (input$id == "FALSE") plotlyOutput("plot")
    
    else if (input$id == "TRUE") dataTableOutput("table")
  }
  else if (input$ipw=="IPW Model"){
    
    if (input$id == "FALSE") plotlyOutput("plot")
    
    else if (input$id == "TRUE") dataTableOutput("table")
  }
})

output$plot <- renderPlotly({
  
  if(input$ipw=="Naive Model"){
  ggplot(malaria_df) +
    aes(x = malaria_risk, fill = net) +
    geom_histogram(bins = 30L) +
    scale_fill_manual(
      values = c(`FALSE` = "#00204D",
      `TRUE` = "#46ADFF")
    ) +
    theme_minimal() ->p
    
  } else if(input$ipw=="MatchIT"){
    
    ggplot(m1data) +
    aes(x = malaria_risk, fill = net) +
    geom_histogram(bins = 30L) +
    scale_fill_manual(
      values = c(`FALSE` = "#00204D",
      `TRUE` = "#46ADFF")
    ) +
    theme_minimal() ->p
    
  } else if(input$ipw == "IPW Model"){
    ggplot(net_ipw, aes(propensity)) +
      geom_mirror_histogram(
        aes(group = net),
        bins = 50
      ) +
      geom_mirror_histogram(
        aes(fill = net, weight = ipw),
        bins = 50,
        alpha = .5
      ) +
      scale_y_continuous(labels = abs) +
      labs(x = "propensity score") +
    scale_fill_manual(
      values = c(`FALSE` = "#00204D",
      `TRUE` = "#46ADFF")
    ) +
    theme_minimal() ->p
  }
    
    ggplotly(p)
  
})

output$table <- renderDT({ 
  
  if(input$ipw=="Naive Model"){
    options(scipen = 99999)
    subset<-malaria_df %>%
      select(net_num,income,temperature,health)
  
  
    output<-tibble()

    cols<-names(subset[2:4])
    
    for(i in  cols){
    
      x<-t.test(subset[[i]][subset$net_num==1],
            subset[[i]][subset$net_num==0])
      
      y<-tidy(x)
      
      z<-tibble(var = i)
      
      t<-cbind(z,y)
      
      output<-bind_rows(output,t)
    
    }
  }    
  
  
  if(input$ipw=="MatchIT"){
    options(scipen = 99999)
    subset<-m1data %>%
      select(net_num,income,temperature,health)
  
  
    output<-tibble()

    cols<-names(subset[2:4])
    
    for(i in  cols){
    
      x<-t.test(subset[[i]][subset$net_num==1],
            subset[[i]][subset$net_num==0], paired = T)
      
      y<-tidy(x)
      
      z<-tibble(var = i)
      
      t<-cbind(z,y)
      
      output<-bind_rows(output,t)
    
    }
  }
  if(input$ipw =="IPW Model"){
    
    options(scipen = 99999)
    subset<-net_ipw %>%
      select(net_num,income,temperature,health, ipw)
    
    output<-tibble()
  
    for(i in names(subset[,2:4])){
      x<- lm(get(i) ~ net_num, subset,weights = ipw)
      
      robust_var <- sandwich(x)[2, 2]
      point_est <- coef(x)[2]
      lb <- point_est - 1.96 * sqrt(robust_var)
      ub <- point_est + 1.96 * sqrt(robust_var)
      model_output <-tibble(
        var = i,
        net = round(point_est,4),
        lower = lb,
        upper = ub,
        "significance at 5% Level" = ifelse((lb >0 & ub> 0) | (lb < 0 & ub < 0) 
                             ,"Significant","Not Significant")
      )
      
      output<-bind_rows(output,model_output)
}
  }
  
  
  
  
  
  
      output
  
  })




```



<b>Difference-in-Difference</b> {data-orientation=rows}
===================================== 



Sidebar {.sidebar}
-------------------------------------

```{r}




radioButtons("id2",label = "Select Plot/Table",
choices = c("Differences","Counterfactual","Data Table"),selected = "Differences")

materialSwitch(inputId = "id3", label = "Show Matched Variables", status = "primary")


```




<h3 class="header">Difference-in-Difference with IPW Matching</h3>

<br>

<p1 class= "p1">Smoking: Dataset on the implementation of Proposition 99 in California in 1988. Data contains information on California and 38 other (control/donor)</p1>


<p2 class= "p2">Proposition 99 increased the cigarette taxes by 0.25$. Also, around this time were anti-smoking campaigns and education on the health risks of smoking.

Counterfactual: <em>what it would have occurred to Y, had the policy intervention not happened?</em></p1>




Row  {data-height=600}
-------------------------------------

### <b>Counterfactual</b>

```{r}

uiOutput("dynamic2")

output$dynamic2 <- renderUI({ 
  if (input$id2 =="Data Table") dataTableOutput("table1")
  else if (input$id2 =="Differences") plotOutput("plot1")
  else if (input$id2 =="Counterfactual") plotOutput("plot2")
})

output$table1 <- renderDT({ 
  
  main
  
})


output$plot1 <- renderPlot({ 
  
  par(omi = c(1, 1, 0, 1)) 
  par(mai = c(.5,1,.5,1))
   
  plot.new()
  min_year<-min(main_plot$year)
  max_year<-max(main_plot$year)
  
  min_cig<-round(min(main_plot$cigs))
  max_cig<-round(max(main_plot$cigs))
  
  plot.window(xlim = c(min_year,max_year), 
              ylim = c(min_cig,max_cig)) 
  
  
  index <- which(main$year >= min_year & main$year <= max_year)
  
  main<-main[index,]
  
  cigs<-main$cigsale
  year<-main$year
  
  control_year<-main_plot$year[main_plot$group==0]
  
  control_cigs<-main_plot$cigs[main_plot$group==0]
  
  
  
  treatment_year<-main_plot$year[main_plot$group==1]
  
  treatment_cigs<-main_plot$cigs[main_plot$group==1]
  
  points(x = year, 
         y = cigs,
         col = "#dddddd",              # Color
         pch = 16,                    # Shape
         cex = 0.75)  
  
  
  
  axis(side = 1,                      # Add x-axis
       lwd = 2,                       # Line width
       hadj = -0.1,                   # Horizontal label position
       padj = -1,                     # Vertical label position
       at = seq(from = min_year,
                to = max_year,
                by = 1))
  
  axis( side = 1, at=c(1987, 1994), 
        padj = 5,
        labels=c("Pre-intervention",
        "Post-intervention" ) )
  
  
  axis(side = 4,                      # Add y-axis
       lwd = 0,                       # Remove axis line
       las = 1,                       # Rotate labels
       col.axis = "black",           # Adjust label colors
       at = seq(from = min_cig,             # Specify axis values
                to = max_cig, 
                by = 5), 
       font.axis = 2)  # Set labels to bold
  
  
  
  lines(control_year, control_cigs, col = "#00204D", pch = 20, cex = 1, type = "o", lwd = 2)
  lines(treatment_year,treatment_cigs, col = "firebrick", pch = 20, cex = 1, type = "o", lwd = 2)
  abline(v=1988, col = "black",pch = 20, cex = 1, lwd = 2)
  segments( x0=1984, y0=95, x1=1985, y1=95, col="#00204D", lwd = 3 )
  points( x=c(1984,1985), y=c(95,95), col="#00204D", pch=20, cex=1.8 )
  text(x = 1986, y = 96, labels = substitute(paste(bold('Control'))), cex = 1.5)
  segments( x0=1984, y0=90, x1=1985, y1=90, col="firebrick", lwd = 3)
  points( x=c(1984,1985), y=c(90,90), col="firebrick", pch=20, cex=1.8 )
  text(x = 1986, y = 90, labels = substitute(paste(bold('Treatment'))), cex = 1.5)
  
  
})


output$plot2<-renderPlot({
  
  
  b0<-coef(x_cig)[1]
  b1<-coef(x_cig)[2]
  b2<-coef(x_cig)[3]
  b3<-coef(x_cig)[4]
  
  
  T1 <- round(b0+b1)
  T2 <- round(b0+b1+b2+b3)
  CF <- round(b0+b1+b2)
  C1 <- round(b0)
  C2 <- round(b0+b2)
  
  palette( c( adjustcolor( "#00204D", alpha.f=1), adjustcolor( "firebrick", alpha.f=1)  ) )
  
  plot( c(0,1,0,1), c(T1,T2,C1,C2), bty="n",
        xlim=c(-1,2), ylim=c(0,135),
        col=c("#00204D","#00204D","firebrick","firebrick"), pch=19, cex=6, 
        xaxt="n", yaxt="n",
        xlab="", ylab="" 
        #main = bquote("Diff-in-Diff" ~ beta[3]), cex.main = 2,
        )
  
  segments( x0=0.08, x1=0.902, y0=C1, y1=C2, col="gray", lwd=2 )
  segments( x0=0.08, x1=0.902, y0=T1, y1=T2, col="gray", lwd=2 )
  
  text( c(0,1,0,1), c(T1,T2,C1,C2), c(T1,T2,C1,C2), cex=1.5, col="white" )
  
  axis( side=1, at=c(0,1), 
        labels=c("Pre-intervention",
        "Post-intervention" ) )
  
  text( 1.1, T2+9, "Treatment Group", col="#00204D", pos=4, cex=1 )
  text( 1.1, C2+9, "Comparison Group", col="firebrick", pos=4, cex=1 )
  
  text( -0.1, T1, expression(paste(beta[0] + beta[1])), col="#00204D", pos=2, cex = 2 )
  text( 1.1, T2, expression(paste(beta[0] + beta[1] + beta[2] + beta[3])), col="#00204D", pos=4, cex=2 )
  text( -0.1, C1, expression(paste(beta[0])), col="firebrick", pos=2, cex = 2 )
  text( 1.1, C2, expression(paste(beta[0] + beta[2])), col="firebrick", pos=4, cex=2 )

  
  cf.col <- "darkgreen"
  
  points( 1, CF, cex=6, col=cf.col )
  segments( x0=0.1, x1=0.9, y0=T1+1, y1=CF-1, col=cf.col, lwd=1, lty=2 )
  text( 1, CF, CF, col=cf.col, cex=1.5 )
  text( 1.1, CF+9, "Counterfactual", col=cf.col, pos=4, cex=1 )
  text( 1.1, CF, expression(paste(beta[0] + beta[1] + beta[2])), col=cf.col, pos=4, cex=2 )
  
})


```




Row  {data-height=400}
-------------------------------------


### <b>Model</b>

```{r}

renderDT({
  
  if (input$id3 ==FALSE){
    
    model_output_cig->p
    
  }
  if(input$id3 ==TRUE){
    output_two<-tibble()
  
    for(i in names(smoking_ipw[,5:7])){
      x<- lm(get(i) ~ group, smoking_ipw,weights = ipw)
      
      robust_var <- sandwich(x)[2, 2]
      point_est <- coef(x)[2]
      lb <- point_est - 1.96 * sqrt(robust_var)
      ub <- point_est + 1.96 * sqrt(robust_var)
      model_output2 <-tibble(
        var = i,
        estimate = point_est,
        lower = lb,
        upper = ub,
        significant = ifelse((lb >0 & ub> 0) | (lb < 0 & ub < 0) 
                             ,"Significant","Not Significant"))
      
      output_two<-bind_rows(output_two,model_output2)->p
      
      
    }
  }
  
  p
  
})






```




```{css,echo = FALSE}

.header{
  text-align:center;
  color: #ffffff;
  font-family: Arial, Helvetica, sans-serif;
}

.p1{
  text-align: left;
  color: #ffffff;
  font-family: Arial, Helvetica, sans-serif;
}

.p2{
  text-align: left;
  color: #ffffff;
  font-family: Arial, Helvetica, sans-serif;
}

.section.sidebar {
    top: 51px;
    background-color: #00204D;
    color:#ffffff;
}


.chart-title {
    border-bottom: 1px solid #dfdfdf;
    font-size: 14px;
    font-weight: 300;
    padding: 7px 10px 4px;
    color: white;
    font-family: inherit;
    background-color: #00204D;
}

```

